Regularising Lotka--Volterra interaction matrices by species grouping: optimal groupings are unpredictable

### 2022-03-20 article.rnw
Fix a missed semantic newline.
Remove the comparison with phylogeny from the article.
The comaprison was based on a qualitative, not quantitative, phylogeny.
As such it was weak, and non-rigorous.

\import{latex/}{forest-phylogeny}

@@

I propose that species might be best partitioned into groups without regard to any \emph{a priori} definition but instead based purely upon which grouping gives the most parsimonious model.
% DBS: We can say something here about how this data-informed approach has worked very well for other clustering problems, such as modularity in networks. I can point to some very good references along those lines.
In this study, I investigate the usefulness and correctness of such an approach using experimental \emph{Drosophila} fecundity data provided by \textcite{terry2021natural}. I fit the Lotka--Volterra model, with grouped species, to find the grouping that gives the most parsimonious model.
I group species independently with regard to their effects upon other species and their responses to other species, to find whether the appropriate groupings might differ for these two cases.
Finally, I compare the resultant groupings with species phylogeny, to provide further investigation of the suitability of \emph{a priori} phylogenetic grouping.

@@

\subsection{Phylogeny}
% DBS: For now, I would cut out the bit about phylogeny since it's relatively weak without a properly calibrated tree. If you want to do a post hoc test, I would go with one that looks at body size and co-classification matrices. On the other hand, my gut tells me that the paper stands on its own merits without a post hoc test, and this is an easy thing to guide reviewers toward commenting on/requesting that you also know how to address. So it might be a useful strategic omission.

I determined the phylogenetic relationships between the six study species by examination of the literature, in order to compare these to the groupings generated by the models.
No attempt was made to determine absolute phylogenetic distances, for sufficient data was not available on all six species for this purpose.
Rather, I only determined relative times of evolutionary divergence sufficient to form a binary phylogenetic tree.

\Textcite{terry2021natural} gave the subgroups of the genus to which each study species belongs.
\Textcite{lewis2005phylogeny} gave the relationship between three of these subgroups (melanogaster, ananassae, and montium), all of the subgenus Sophophora, while \textcite{izumitani2016phylogeography} described the fourth subgroup (nasuta) as belonging to the distinct Drosophila subgenus.
From this can be constructed the necessary relationships, which are depicted in \cref{fig:phylogeny}.

\begin{figure}
	\centering
	\begin{forest} phylogeny
		[
			[
				[
					[\emph{D. simulans}]
					[\emph{D. birchii}]
				]
				[
					[\emph{D. pandora}]
					[\emph{D. pseudoananassae}]
				]
			]
			[
				[
					[\emph{D. pallidifrons}]
					[\emph{D. sulfurigaster}]
				]
			]
		]
	\end{forest}
	%TODO: Mark subgroups and subgenera on the diagram.

	\caption{%
		The phylogenetic relationships between the six \emph{Drosophila} species in the study.
		The relationships depicted are purely qualitative; no implication of phylogenetic distance is made.
	}
	\label{fig:phylogeny}
\end{figure}

@@

A high likelihood of coclassification was associated with phylogeny in one instance, but by no means consistently.
The phylogeny divided the six species into three pairs of most closely related species (\cref{fig:phylogeny}).
The pair of \emph{D. pallidifrons} and \emph{D. sulfurigaster}, which both belonged to a different subgenus than the other four species, were also the only pair of species to be coclassified with relatively high probability in both the row and column groupings.
In contrast, both of the other most closely related pairs were coclassified with very low probabilities in both the row and column groupings.

@@

I chose a study system of \emph{Drosophila} partially on the basis that it would allow clear analysis of the species' phylogenetic relationships.
Phylogeny proved a poor predictor of optimal grouping, as previous studies have suggested that it might \parencite{godoy2014phylogenetic,cahill2008does}.
A study system with species spanning multiple functional groups, however, might have shown clearer groupings.
While functional groups have previously shown to be poor predictors of competitive interactions \parencite{bret2008plant}, studies in other contexts have succeeded in isolating functional groups by selecting for model parsimony \parencite{allesina2009food}.
In further support of functional groups over phylogeny, \textcite{uriarte2004spatially} found that grouping based on light guilds (akin to functional groups; \cite{blondel2003guilds}) produced a parsimonious model more frequently than grouping taxonomically (with both groupings defined \emph{a priori}).
Perhaps an attempt to produce functional groups based upon competitive interactions is warranted, applying the framework I present here to a set of species spanning multiple functional groups.
% DBS: Prolix warning! ;)

@@

All in all, the method of \emph{a posteriori} parsimony-based selection of groups has not proven to be greatly effective, at least on such a closely related set of species.
However, the results suggest a number of pieces of advice which might be borne in mind whenever selecting groupings \emph{a priori} for use in species interaction models.
Firstly, there may be no single best grouping, but perhaps multiple rather distinct groupings which each perform similarly.
Secondly, phylogeny does not appear to provide appropriate groupings.
Lastly, one should not implicitly assume that species are grouped in the same way with regard to their effects upon other species and their response to them; distinct groupings of the two might well be considered.

>>>>

@@

I propose that species might be best partitioned into groups without regard to any \emph{a priori} definition but instead based purely upon which grouping gives the most parsimonious model.
% DBS: We can say something here about how this data-informed approach has worked very well for other clustering problems, such as modularity in networks. I can point to some very good references along those lines.
In this study, I investigate the usefulness and correctness of such an approach using experimental \emph{Drosophila} fecundity data provided by \textcite{terry2021natural}.
I fit the Lotka--Volterra model, with grouped species, to find the grouping that gives the most parsimonious model.
I group species independently with regard to their effects upon other species and their responses to other species, to find whether the appropriate groupings might differ for these two cases.

@@

@@

@@

%TODO: Discuss the potential to study a system with functional groups without brining up phylogeny.
%I chose a study system of \emph{Drosophila} partially on the basis that it would allow clear analysis of the species' phylogenetic relationships.
%Phylogeny proved a poor predictor of optimal grouping, as previous studies have suggested that it might \parencite{godoy2014phylogenetic,cahill2008does}.
%A study system with species spanning multiple functional groups, however, might have shown clearer groupings.
%While functional groups have previously shown to be poor predictors of competitive interactions \parencite{bret2008plant}, studies in other contexts have succeeded in isolating functional groups by selecting for model parsimony \parencite{allesina2009food}.
%In further support of functional groups over phylogeny, \textcite{uriarte2004spatially} found that grouping based on light guilds (akin to functional groups; \cite{blondel2003guilds}) produced a parsimonious model more frequently than grouping taxonomically (with both groupings defined \emph{a priori}).
%Perhaps an attempt to produce functional groups based upon competitive interactions is warranted, applying the framework I present here to a set of species spanning multiple functional groups.
% DBS: Prolix warning! ;)

@@

All in all, the method of \emph{a posteriori} parsimony-based selection of groups has not proven to be greatly effective, at least on such a closely related set of species.
However, the results suggest some advice which might be borne in mind whenever selecting groupings \emph{a priori} for use in species interaction models.
Firstly, there may be no single best grouping, but perhaps multiple rather distinct groupings which each perform similarly.
Furthermore, one should not implicitly assume that species are grouped in the same way with regard to their effects upon other species and their response to them; distinct groupings of the two might well be considered.
